---
title: "IPUMS data collection and processing"
author: "Harrison DeFord"
date: '2022-11-18'
output: html_document
---

This document describes the process through which population estimates were made and related to water area values extracted from Google Earth.

```{r message = FALSE}
library(ipumsr)
library(dplyr)
library(stringr)
library(tidycensus)
library(ggplot2)
library(gganimate)
library(tidyverse)
```

First, a CSV containing information from the 1980, 1990, 2000, and 2010 censuses, all normalized to 2010 geographies, is loaded using `read_nhgis()` from `{ipumsr}`, and only counties bordering the Great Salt Lake were kept.

```{r warning = FALSE, message=FALSE, comment=FALSE}
ut_ts <- read_nhgis("../data/nhgis0005_ts_geog2010_tract.csv") %>%
  filter(STATE == "Utah") %>%
  filter(str_detect(COUNTY, "Utah County|Salt Lake County|Davis|Weber|Box Elder|Tooele")) %>%
  select(c(1,6,8:17)) %>%
  pivot_longer(cols = tidyselect::starts_with("CL8AA"))%>%
  filter(name %in% c("CL8AA1990","CL8AA2000","CL8AA2010","CL8AA2020"))
```

A linear regression model was generated for each census tract. This allows for the interpolation of population data between the 10-year census recurrence interval, before the ACS was begun in 2009.

```{r include=FALSE}
ut_ts.lm <- ut_ts %>% mutate(year = str_sub(ut_ts$name, start = -4,-1), value = round(value)) %>%
  select(-name) %>%
  filter(as.numeric(year) <= 2010)
ut_ts.split <- ut_ts.lm %>%
  group_by(GISJOIN) %>%
  group_split()
ut_model <- lapply(ut_ts.split, function(x) lm(value~as.numeric(year), x))
summary(ut_model)
```

The estimated annual populations of each tract were summarized to the level of the entire geography, then ACS 5-year data was used from 2009 onwards. These data sources were combined to yield continuous annual population estimates from 1986 to 2020.

```{r message=FALSE}
ut_ts.pred <- list()
for (i in 1:length(ut_model)){
  ut_ts.pred[[i]] = data.frame(year = c(1986:2009), 
                               GISJOIN = first(ut_ts.split[[i]]$GISJOIN), 
                               COUNTYA = first(ut_ts.split[[i]]$COUNTYA))
  ut_ts.pred[[i]]$tpop = round(predict(ut_model[[i]], newdata = ut_ts.pred[[i]]))
} 
ut_ts.pred <- bind_rows(ut_ts.pred)

ut_ts.sum <- ut_ts.pred %>%
  group_by(year)%>%
  mutate(tpop = sum(tpop)) %>%
  ungroup()%>%
  distinct(year, .keep_all = TRUE) %>%
  select(year, tpop)

ut_acs.list <- lapply(c(2010:2020), function(y) get_acs(geography = "county", state = "UT", year = y, survey = "acs5", variables = "B01001_001", cache_table = TRUE) %>% mutate(year = y)) %>%
  bind_rows()
ut_acs.filtered <- ut_acs.list %>% filter(str_detect(NAME, "Utah County|Salt Lake County|Davis|Weber|Box Elder|Tooele")) %>%
  group_by(year) %>%
  mutate(tpop = sum(estimate))%>%
  ungroup()%>%
  distinct(year, .keep_all = TRUE) %>%
  select(year, tpop)

ut_tpop <- bind_rows(ut_ts.sum, ut_acs.filtered)
```

Finally, the population estimates were joined to the water area values obtained from Google Earth Engine, and written to a .csv for use in further processing.

```{r eval=FALSE}
year_area_gsl <- read_csv("../data/year_area_gsl.csv")
vars_joined <- left_join(ut_tpop, year_area_gsl) %>% mutate(area_km = (area/1e6), tpop_mn = (tpop/1e6))
vars_joined <- na.omit(vars_joined)
write_csv(vars_joined, file = "../data/vars_joined.csv")
```
