---
title: "scooter analysis"
author: "Harrison DeFord"
date: '2022-05-05'
output: html_document
execute: 
  eval: false
---

```{r setup, include=FALSE}
library(geojsonsf)
library(jsonlite)
library(httr)
library(sf)
library(tibble)
library(lubridate)
library(tidycensus)
library(tigris)
library(stringr)
library(dplyr)
library(tmap)
library(mapdeck)
tmap_options(check.and.fix = TRUE)
knitr::opts_chunk$set(echo = TRUE)
options(tigris_use_cache = TRUE)
```

## Setup

This script relies on 3 geopackage outputs from the previous script. It creates a long dataframe, inclusive of all three vendors, groups them by permanent, unique bike ID, and defines a trip as a movement of 50 meters between timestamps, in an attempt to adjust for GPS variability.

```{r read_vars, message = FALSE}
link_data <- st_read("../results/link_mon_am.gpkg")
lime_data <- st_read("../results/spin_mon_am.gpkg")
spin_data <- st_read("../results/spin_mon_am.gpkg")
scooters_raw = bind_rows(link_data, lime_data, spin_data) %>%
  st_transform(crs = 3857)
if(!file.exists("../results/scooters_raw.gpkg")){
  st_write(scooters_raw, dsn = "../results/scooters_raw.gpkg")
}
```

First, we'll filter out the disabled bike observations, and split the bikes into their own lists in order to apply a function over each bike.

```{r group_split, message = FALSE}
scooters_split <- scooters_raw %>%
  distinct() %>%
  filter(is_disabled == 0) %>%
  group_by(vendor, bike_id) %>%
  group_split()
```

This function defines what constitutes a trip. It is *slow*. When it's done, each list element, representing a scooter, will have a field delineating if a given time-interval was part of a trip.

```{r is_trip, message = FALSE}
scooters_trip <- lapply(scooters_split, function(df){
  df %>% mutate(dist_prev = units::drop_units(st_distance(geom, lag(geom), by_element = TRUE)),
                dist_next = units::drop_units(st_distance(geom, lead(geom), by_element = TRUE)),
                time_id = row_number(), #this is what allows us to order points for QGIS analysis
                movement_id = paste(bike_id, "_", row_number(), sep = ""), #perhaps redundant, but easy solution for moving between R and QGIS
                trip = case_when(
                  dist_prev > 50 | dist_next > 50 ~ 1, #define trip based on distance column
                  TRUE ~ 0))
})
```

Note that this is some of the earliest R code I'd ever written. I'm leaving it alone for posterity's sake, but I'm not sure why I split it into groups and re-bound it again.

```{r filter_trip, message = FALSE}
trip_long <- bind_rows(scooters_trip) %>%
  filter(trip == 1) #filter by only trip points
trip_split <- trip_long %>% #split again by trips
  group_by(bike_id) %>%
  group_split()
```

```{r write, message = FALSE}
#trip_split_id <- lapply(trip_split, function(df){
#  df %>% mutate(time_id = row_number())
#})
trip_id_long <-bind_rows(trip_split) #bind into trips
if(!file.exists("../results/trip_id_long.gpkg")){
  st_write(trip_id_long, "../results/trip_id_long.gpkg", append = FALSE)
}
```
